---
title: "Attributing Pneumonia Hospitalizations to Respiratory Viruses"
subtitle: "A Time Series Decomposition Framework"
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  fig-width: 11
  fig-height: 6
---

```{r setup}
library(tidyverse)
library(patchwork)
library(scales)

set.seed(2024)

# ── Consistent color palette across all figures ───────────────────────────
pal <- c(
  baseline = "#8B9DB5",
  flu      = "#E55C30",
  rsv      = "#27AE60",
  covid    = "#A93226"
)

# ── Shared ggplot theme ───────────────────────────────────────────────────
theme_ts <- function(...) {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor   = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title.x       = element_blank(),
      legend.position    = "bottom",
      legend.title       = element_blank(),
      plot.title         = element_text(face = "bold", size = 13),
      plot.subtitle      = element_text(color = "gray40", size = 10)
    ) +
    theme(...)
}
```

---

## Data

```{r simulate-data}
# (Pandemic-era data not used — all analyses use pre-pandemic df2 below)
```

---

## Pre-Pandemic Data (2010–2019)

```{r fig2-data}
# Pre-pandemic data (2010–2019) — all analyses use this dataset
n2     <- 522  # ~10 years of weekly data
dates2 <- seq(as.Date("2010-01-03"), by = "week", length.out = n2)
t2     <- seq_len(n2)
wk2    <- as.integer(format(dates2, "%U"))

# ── Influenza: year-to-year variation in peak timing and intensity ─────────
# Assign each week to a flu season (Jul Y → Jun Y+1, labelled Y+1)
flu_season_yr <- as.integer(format(dates2, "%Y")) +
                 ifelse(as.integer(format(dates2, "%m")) >= 7L, 1L, 0L)
seasons   <- sort(unique(flu_season_yr))
n_seasons <- length(seasons)

# Each season: random ±3-week peak shift and log-normal amplitude multiplier
peak_offsets <- round(rnorm(n_seasons, 0, 2.5))   # SD ≈ 2–3 weeks
amplitudes   <- exp(rnorm(n_seasons, 0, 0.45))    # ~50 % CV in intensity

flu2 <- numeric(n2)
for (i in seq_along(seasons)) {
  idx  <- which(flu_season_yr == seasons[i])
  pw   <- 2L + peak_offsets[i]               # typical peak week 2 ± shift
  flu2[idx] <- pmax(0, cos(2 * pi * (wk2[idx] - pw) / 52))^5 *
               90 * amplitudes[i]             # ^5 gives sharper winter spike
}
flu2 <- pmax(0, flu2 + rnorm(n2, 0, 4))

# ── RSV: weak biennial variation (higher/lower amplitude in alternate years)
rsv_year    <- as.integer(format(dates2, "%Y"))
biennl_mult <- ifelse(rsv_year %% 2 == 0, 1.40, 0.65)
rsv2 <- pmax(0,
  pmax(0, cos(2 * pi * (wk2 - 47) / 52))^2.5 * 55 * biennl_mult + rnorm(n2, 0, 3)
)

# ── Scale to [0, 1] for model predictors ──────────────────────────────────
flu2_s <- flu2 / max(flu2)
rsv2_s <- rsv2 / max(rsv2)

# ── Harmonic terms ─────────────────────────────────────────────────────────
sin1_2 <- sin(2 * pi * t2 / 52);  cos1_2 <- cos(2 * pi * t2 / 52)
sin2_2 <- sin(4 * pi * t2 / 52);  cos2_2 <- cos(4 * pi * t2 / 52)

# ── Pneumonia: identity link; nonlinear virus interaction ────────────────
# mu = int + seasonal harmonics + exp(b1*flu + b2*rsv) + b3*rsv + b4*flu
int2   <- 200   # baseline count
b3_rsv <- 30    # linear rsv effect (counts)
b4_flu <- 80    # linear flu effect (counts)

mu2 <- (int2 +
          12 * sin1_2 - 8 * cos1_2 +            # seasonal baseline
           4 * sin2_2 + 2 * cos2_2 +
          b3_rsv * rsv2_s + b4_flu * flu2_s)    # linear virus effects
mu2 <- pmax(1, mu2)                             # keep mu positive

pneu2 <- rnbinom(n2, mu = mu2, size = 40)       # size↑ → cleaner signal

df2 <- tibble(
  date  = dates2,
  t     = t2,
  pneu  = pneu2,
  flu   = flu2, rsv = rsv2,
  flu_s = flu2_s, rsv_s = rsv2_s,
  sin1  = sin1_2, cos1 = cos1_2,
  sin2  = sin2_2, cos2 = cos2_2
)

winters2 <- tibble(
  xmin = as.Date(paste0(2009:2018, "-12-01")),
  xmax = as.Date(c(paste0(2010:2018, "-02-28"), "2019-02-28"))
)
```

---

## Figure 1: The Raw Input Data

```{r fig1}
#| fig-cap: "Weekly time series for each variable, 2010–2019 (pre-pandemic). Blue bands mark winter months (December–February), when respiratory viruses tend to peak."
#| fig-height: 7

# Long format: one row per date × series (pre-pandemic data only)
df_long1 <- df2 |>
  select(date, pneu, flu, rsv) |>
  pivot_longer(-date, names_to = "series", values_to = "value") |>
  mutate(
    panel = factor(series,
      levels = c("pneu", "flu", "rsv"),
      labels = c(
        "Pneumonia Hospitalizations  (weekly count)",
        "Influenza Activity  (index, 0 – 100)",
        "RSV Activity  (index, 0 – 100)"
      )
    ),
    color = case_match(series,
      "pneu" ~ "black",
      "flu"  ~ unname(pal["flu"]),
      "rsv"  ~ unname(pal["rsv"])
    )
  )

ggplot(df_long1, aes(date, value)) +
  geom_rect(
    data        = winters2,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill        = "steelblue", alpha = 0.09
  ) +
  geom_line(aes(color = color), linewidth = 0.65, show.legend = FALSE) +
  facet_wrap(~panel, scales = "free_y", ncol = 1) +
  scale_color_identity() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(y = NULL, title = "Figure 1: Raw Input Data  (2010–2019)") +
  theme_ts(
    panel.grid.major.x = element_line(color = "gray90"),
    strip.text         = element_text(face = "bold", size = 11)
  )
```

---

## Model Fitting

```{r fit-model}
# ── Fit Poisson GLM on pre-pandemic data (no COVID term) ──────────────────
fit <- glm(
  pneu ~ sin1 + cos1 + sin2 + cos2 + flu_s + rsv_s,
  data   = df2,
  family = poisson(link = "log")
)

# Fitted values and 95% pointwise CI on the mean
pred <- predict(fit, se.fit = TRUE)
df2 <- df2 |>
  mutate(
    fitted    = exp(pred$fit),
    fit_lo    = exp(pred$fit - 1.96 * pred$se.fit),
    fit_hi    = exp(pred$fit + 1.96 * pred$se.fit),
    pearson_r = residuals(fit, type = "pearson")
  )

# ── Decompose fitted values into additive components ─────────────────────
b <- coef(fit)

lp_base     <- b["(Intercept)"]
lp_seasonal <- b["sin1"]*df2$sin1 + b["cos1"]*df2$cos1 +
               b["sin2"]*df2$sin2 + b["cos2"]*df2$cos2
lp_flu      <- b["flu_s"] * df2$flu_s
lp_rsv      <- b["rsv_s"] * df2$rsv_s

mu_none <- exp(lp_base + lp_seasonal)                  # no viruses
mu_rsv  <- exp(lp_base + lp_seasonal + lp_rsv)         # +RSV
mu_full <- df2$fitted                                   # +RSV +flu

df2 <- df2 |>
  mutate(
    comp_baseline = mu_none,
    comp_rsv      = pmax(0, mu_rsv  - mu_none),
    comp_flu      = pmax(0, mu_full - mu_rsv)
  )
```

---

## Link Function Comparison: Log vs Identity

```{r link-comparison}
# ── Fit both link functions ────────────────────────────────────────────────
fit_log <- fit  # already fitted above

fit_id <- glm(
  pneu ~ sin1 + cos1 + sin2 + cos2 + flu_s + rsv_s,
  data    = df2,
  family  = poisson(link = "identity"),
  start   = coef(fit_log) |> (\(b) c(exp(b[1]), b[-1]))()  # sensible starting values
)

# ── Collect GOF statistics ─────────────────────────────────────────────────
gof <- tibble(
  Link         = c("log", "identity"),
  AIC          = c(AIC(fit_log), AIC(fit_id)),
  Deviance     = c(deviance(fit_log), deviance(fit_id)),
  Null_Dev     = c(fit_log$null.deviance, fit_id$null.deviance),
  df_resid     = c(df.residual(fit_log), df.residual(fit_id))
) |>
  mutate(
    Pseudo_R2    = 1 - Deviance / Null_Dev,
    RMSE         = c(
      sqrt(mean((df2$pneu - fitted(fit_log))^2)),
      sqrt(mean((df2$pneu - fitted(fit_id))^2))
    )
  ) |>
  select(Link, AIC, Deviance, Pseudo_R2, RMSE)

knitr::kable(
  gof,
  digits  = c(0, 1, 1, 4, 2),
  caption = "Table 1: Goodness-of-fit comparison — Poisson log vs identity link",
  col.names = c("Link", "AIC", "Residual Deviance", "Pseudo R²", "RMSE")
)
```

```{r link-comparison-plot}
#| fig-cap: "Observed weekly pneumonia counts (black points) vs fitted values from Poisson models with log link (blue) and identity link (orange). Shaded bands are 95% pointwise CIs on the mean."

pred_id <- predict(fit_id, se.fit = TRUE)

df_compare <- df2 |>
  select(date, pneu) |>
  mutate(
    log_fit    = fitted(fit_log),
    log_lo     = exp(predict(fit_log, se.fit = TRUE)$fit - 1.96 * predict(fit_log, se.fit = TRUE)$se.fit),
    log_hi     = exp(predict(fit_log, se.fit = TRUE)$fit + 1.96 * predict(fit_log, se.fit = TRUE)$se.fit),
    id_fit     = fitted(fit_id),
    id_lo      = pmax(0, pred_id$fit - 1.96 * pred_id$se.fit),
    id_hi      = pred_id$fit + 1.96 * pred_id$se.fit
  )

ggplot(df_compare, aes(x = date)) +
  geom_ribbon(aes(ymin = log_lo, ymax = log_hi), fill = "#2166ac", alpha = 0.15) +
  geom_ribbon(aes(ymin = id_lo,  ymax = id_hi),  fill = "#d6604d", alpha = 0.15) +
  geom_line(aes(y = log_fit, color = "log"),      linewidth = 0.7) +
  geom_line(aes(y = id_fit,  color = "identity"), linewidth = 0.7, linetype = "dashed") +
  geom_point(aes(y = pneu), color = "black", size = 0.6, alpha = 0.45) +
  scale_color_manual(
    values = c("log" = "#2166ac", "identity" = "#d6604d"),
    labels = c("log" = "Poisson (log)", "identity" = "Poisson (identity)")
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma_format()) +
  labs(
    y       = "Weekly hospitalizations",
    color   = "Link function",
    title   = "Figure 2: Fitted Values by Link Function",
    subtitle = "Shaded bands = 95% CI on fitted mean"
  ) +
  theme_ts(legend.position = "top")
```

```{r burden-comparison}
#| fig-cap: "10-year attributable burden estimates (RSV and Influenza) from both link functions, with percentage of total modeled hospitalizations."

# ── Decompose identity-link model ─────────────────────────────────────────
# Identity link: mu = b0 + b_seas + b_flu*flu + b_rsv*rsv
# so components are additive directly — no exp() needed
b_id <- coef(fit_id)

id_baseline <- b_id["(Intercept)"] +
               b_id["sin1"]*df2$sin1 + b_id["cos1"]*df2$cos1 +
               b_id["sin2"]*df2$sin2 + b_id["cos2"]*df2$cos2

id_comp_rsv <- pmax(0, b_id["rsv_s"] * df2$rsv_s)
id_comp_flu <- pmax(0, b_id["flu_s"] * df2$flu_s)

# ── True attributable burden from the data-generating process ─────────────
# DGP is purely linear: mu = int + seasonal + b3_rsv*rsv_s + b4_flu*flu_s
# so true component contributions are simply the linear terms summed over time
true_burden <- tibble(
  Pathogen   = c("RSV", "Influenza"),
  TrueBurden = c(
    sum(b3_rsv * df2$rsv_s),
    sum(b4_flu * df2$flu_s)
  )
)

# ── Summary table ─────────────────────────────────────────────────────────
burden <- bind_rows(
  tibble(
    Link      = "log",
    Pathogen  = c("RSV", "Influenza"),
    Burden    = c(sum(df2$comp_rsv), sum(df2$comp_flu)),
    Total_fit = sum(df2$fitted)
  ),
  tibble(
    Link      = "identity",
    Pathogen  = c("RSV", "Influenza"),
    Burden    = c(sum(id_comp_rsv), sum(id_comp_flu)),
    Total_fit = sum(fitted(fit_id))
  )
) |>
  mutate(
    Pct        = Burden / Total_fit * 100,
    bar_label  = sprintf("%s  (%.1f%%)", comma(round(Burden)), Pct)
  )

knitr::kable(
  burden |> select(Link, Pathogen, Burden, Pct),
  digits    = c(0, 0, 0, 1),
  caption   = "Table 2: 10-year attributable burden by link function",
  col.names = c("Link", "Pathogen", "Attributable hospitalizations", "% of total fitted")
)
```

```{r burden-comparison-plot}
#| fig-cap: "Side-by-side comparison of 10-year attributable burden for RSV and Influenza under Poisson log vs identity link. Percentages are of total modeled hospitalizations under each model."
#| fig-height: 3.5

ggplot(burden, aes(x = Pathogen, y = Burden, fill = Link)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  geom_text(
    aes(label = bar_label),
    position = position_dodge(width = 0.6),
    vjust    = -0.4,
    size     = 3.2,
    fontface = "bold",
    color    = "gray20"
  ) +
  geom_crossbar(
    data        = true_burden,
    aes(x = Pathogen, y = TrueBurden, ymin = TrueBurden, ymax = TrueBurden),
    inherit.aes = FALSE,
    width       = 0.65,
    color       = "black",
    linewidth   = 0.7,
    linetype    = "dashed",
    fatten      = 0
  ) +
  scale_fill_manual(
    values = c("log" = "#2166ac", "identity" = "#d6604d"),
    labels = c("log" = "Poisson (log)", "identity" = "Poisson (identity)")
  ) +
  scale_y_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    x        = NULL,
    y        = "Attributable hospitalizations (10-year total)",
    fill     = NULL,
    title    = "Figure 3: Attributable Burden — Log vs Identity Link",
    subtitle = "Excess above seasonal baseline, by pathogen and link function\nDashed line = true value from data-generating process"
  ) +
  theme_ts(legend.position = "top")
```

---

## Figure 5: Decomposing the Signal

```{r fig5}
#| fig-cap: "Stacked area chart showing the contribution of each model component to weekly expected hospitalizations. The total stack matches the model's fitted values at every time point. Observed counts are overlaid as black points."

decomp_long <- df2 |>
  select(date, pneu, comp_baseline, comp_rsv, comp_flu) |>
  pivot_longer(
    cols      = starts_with("comp_"),
    names_to  = "component",
    values_to = "value"
  ) |>
  mutate(component = factor(component,
    levels = c("comp_baseline", "comp_rsv", "comp_flu"),
    labels = c("Seasonal Baseline", "RSV", "Influenza")
  ))

comp_colors <- c(
  "Seasonal Baseline" = unname(pal["baseline"]),
  "RSV"               = unname(pal["rsv"]),
  "Influenza"         = unname(pal["flu"])
)

ggplot(decomp_long, aes(date, value, fill = component)) +
  geom_rect(
    data        = winters2,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill        = "white", alpha = 0.35
  ) +
  geom_area(position = "stack", alpha = 0.88, color = NA) +
  geom_point(
    data        = df2,
    aes(x = date, y = pneu),
    inherit.aes = FALSE,
    color       = "black", size = 0.75, alpha = 0.50
  ) +
  scale_fill_manual(values = comp_colors) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma_format()) +
  labs(
    y        = "Expected hospitalizations / week",
    title    = "Figure 5: Decomposing Pneumonia Hospitalizations by Component",
    subtitle = "Stacked areas = model components; points = observed counts",
    fill     = NULL
  ) +
  theme_ts() +
  guides(fill = guide_legend(nrow = 1, reverse = TRUE))
```

---

## Figure 6: Attributable Burden Summary

```{r fig6}
#| fig-cap: "Estimated hospitalizations attributable to each respiratory virus over the full 10-year pre-pandemic period (excess above the seasonal baseline). Percentages are of total modeled hospitalizations."
#| fig-height: 3.8

virus_totals <- tibble(
  Component = c("Influenza", "RSV"),
  Total     = c(sum(df2$comp_flu), sum(df2$comp_rsv)),
  Color     = c(pal["flu"], pal["rsv"])
) |>
  mutate(
    pct_total = Total / sum(df2$fitted) * 100,
    bar_label = sprintf(
      "%s hospitalizations  (%.1f%% of total)",
      comma(round(Total)),
      pct_total
    ),
    Component = fct_reorder(Component, Total)
  )

ggplot(virus_totals, aes(x = Component, y = Total, fill = Color)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(
    aes(label = bar_label),
    hjust  = -0.05,
    size   = 3.8,
    fontface = "bold",
    color  = "gray20"
  ) +
  coord_flip(clip = "off") +
  scale_fill_identity() +
  scale_y_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.55))
  ) +
  labs(
    x        = NULL,
    y        = "Estimated attributable hospitalizations  (10-year total)",
    title    = "Figure 6: Virus-Attributable Hospitalization Burden",
    subtitle = "Excess hospitalizations above seasonal baseline, by pathogen"
  ) +
  theme_ts(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "gray88"),
    axis.text.y        = element_text(size = 12, face = "bold"),
    plot.margin        = margin(10, 120, 10, 10)
  )
```

---

## Multiplicative DGP: Simulation & Model Comparison

The analyses above used an **additive** data-generating process (identity link is the
"correct" model). Here we repeat the exercise with a **multiplicative** DGP —
`mu = exp(intercept + seasonal + b_rsv · rsv + b_flu · flu)` — so the Poisson **log**
link is now correctly specified, and the identity link is misspecified. Comparing the
two scenarios shows how link-function misspecification affects burden estimation in each
direction.

```{r mult-simulate}
# ── Multiplicative DGP parameters (log scale) ─────────────────────────────
int2_m   <- log(200)   # log intercept → exp ≈ 200 baseline counts
b3_rsv_m <- 0.15       # log-rate RSV effect  (≈+16% at max RSV)
b4_flu_m <- 0.40       # log-rate flu effect  (≈+49% at max flu)

mu2_m <- exp(
  int2_m +
  12 * sin1_2 - 8 * cos1_2 +       # seasonal baseline
   4 * sin2_2 + 2 * cos2_2 +
  b3_rsv_m * rsv2_s + b4_flu_m * flu2_s
)

pneu2_m <- rnbinom(n2, mu = mu2_m, size = 40)

df2_m <- tibble(
  date  = dates2,
  t     = t2,
  pneu  = pneu2_m,
  flu_s = flu2_s, rsv_s = rsv2_s,
  sin1  = sin1_2, cos1 = cos1_2,
  sin2  = sin2_2, cos2 = cos2_2
)
```

```{r mult-fit}
# ── Fit Poisson log link ───────────────────────────────────────────────────
fit_log_m <- glm(
  pneu ~ sin1 + cos1 + sin2 + cos2 + flu_s + rsv_s,
  data   = df2_m,
  family = poisson(link = "log")
)

# ── Fit Poisson identity link ──────────────────────────────────────────────
fit_id_m <- glm(
  pneu ~ sin1 + cos1 + sin2 + cos2 + flu_s + rsv_s,
  data   = df2_m,
  family = poisson(link = "identity"),
  start  = coef(fit_log_m) |> (\(b) c(exp(b[1]), b[-1]))()
)

# ── GOF table ──────────────────────────────────────────────────────────────
gof_m <- tibble(
  Link     = c("log", "identity"),
  AIC      = c(AIC(fit_log_m), AIC(fit_id_m)),
  Deviance = c(deviance(fit_log_m), deviance(fit_id_m)),
  Null_Dev = c(fit_log_m$null.deviance, fit_id_m$null.deviance)
) |>
  mutate(
    Pseudo_R2 = 1 - Deviance / Null_Dev,
    RMSE      = c(
      sqrt(mean((df2_m$pneu - fitted(fit_log_m))^2)),
      sqrt(mean((df2_m$pneu - fitted(fit_id_m))^2))
    )
  ) |>
  select(Link, AIC, Deviance, Pseudo_R2, RMSE)

knitr::kable(
  gof_m,
  digits    = c(0, 1, 1, 4, 2),
  caption   = "Table 3: Multiplicative DGP — Poisson log vs identity link",
  col.names = c("Link", "AIC", "Residual Deviance", "Pseudo R²", "RMSE")
)
```

```{r mult-fit-plot}
#| fig-cap: "Multiplicative DGP: observed weekly counts (black points) vs fitted values from the correctly-specified Poisson log-link model (blue) and the misspecified identity-link model (orange). Shaded bands are 95% pointwise CIs."

pred_log_m <- predict(fit_log_m, se.fit = TRUE)
pred_id_m  <- predict(fit_id_m,  se.fit = TRUE)

df_compare_m <- df2_m |>
  select(date, pneu) |>
  mutate(
    log_fit = fitted(fit_log_m),
    log_lo  = exp(pred_log_m$fit - 1.96 * pred_log_m$se.fit),
    log_hi  = exp(pred_log_m$fit + 1.96 * pred_log_m$se.fit),
    id_fit  = fitted(fit_id_m),
    id_lo   = pmax(0, pred_id_m$fit - 1.96 * pred_id_m$se.fit),
    id_hi   = pred_id_m$fit + 1.96 * pred_id_m$se.fit
  )

ggplot(df_compare_m, aes(x = date)) +
  geom_ribbon(aes(ymin = log_lo, ymax = log_hi), fill = "#2166ac", alpha = 0.15) +
  geom_ribbon(aes(ymin = id_lo,  ymax = id_hi),  fill = "#d6604d", alpha = 0.15) +
  geom_line(aes(y = log_fit, color = "log"),      linewidth = 0.7) +
  geom_line(aes(y = id_fit,  color = "identity"), linewidth = 0.7, linetype = "dashed") +
  geom_point(aes(y = pneu), color = "black", size = 0.6, alpha = 0.45) +
  scale_color_manual(
    values = c("log" = "#2166ac", "identity" = "#d6604d"),
    labels = c("log" = "Poisson (log)", "identity" = "Poisson (identity)")
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma_format()) +
  labs(
    y        = "Weekly hospitalizations",
    color    = "Link function",
    title    = "Figure 7: Fitted Values — Multiplicative DGP",
    subtitle = "Shaded bands = 95% CI on fitted mean"
  ) +
  theme_ts(legend.position = "top")
```

```{r mult-burden}
# ── Decompose log-link model (sequential: +RSV then +flu) ─────────────────
b_log_m       <- coef(fit_log_m)
lp_base_m     <- b_log_m["(Intercept)"]
lp_seasonal_m <- b_log_m["sin1"]*df2_m$sin1 + b_log_m["cos1"]*df2_m$cos1 +
                 b_log_m["sin2"]*df2_m$sin2 + b_log_m["cos2"]*df2_m$cos2

mu_none_log_m <- exp(lp_base_m + lp_seasonal_m)
mu_rsv_log_m  <- exp(lp_base_m + lp_seasonal_m + b_log_m["rsv_s"] * df2_m$rsv_s)
mu_full_log_m <- fitted(fit_log_m)

log_m_comp_rsv <- pmax(0, mu_rsv_log_m  - mu_none_log_m)
log_m_comp_flu <- pmax(0, mu_full_log_m - mu_rsv_log_m)

# ── Decompose identity-link model ─────────────────────────────────────────
b_id_m        <- coef(fit_id_m)
id_m_comp_rsv <- pmax(0, b_id_m["rsv_s"] * df2_m$rsv_s)
id_m_comp_flu <- pmax(0, b_id_m["flu_s"] * df2_m$flu_s)

# ── True burden from multiplicative DGP (sequential decomposition) ─────────
lp_seas_true_m <- 12*df2_m$sin1 - 8*df2_m$cos1 + 4*df2_m$sin2 + 2*df2_m$cos2

mu_none_true_m <- exp(int2_m + lp_seas_true_m)
mu_rsv_true_m  <- exp(int2_m + lp_seas_true_m + b3_rsv_m * df2_m$rsv_s)
mu_full_true_m <- exp(int2_m + lp_seas_true_m + b3_rsv_m * df2_m$rsv_s + b4_flu_m * df2_m$flu_s)

true_burden_m <- tibble(
  Pathogen   = c("RSV", "Influenza"),
  TrueBurden = c(
    sum(mu_rsv_true_m  - mu_none_true_m),
    sum(mu_full_true_m - mu_rsv_true_m)
  )
)

# ── Burden summary table ───────────────────────────────────────────────────
burden_m <- bind_rows(
  tibble(
    Link      = "log",
    Pathogen  = c("RSV", "Influenza"),
    Burden    = c(sum(log_m_comp_rsv), sum(log_m_comp_flu)),
    Total_fit = sum(fitted(fit_log_m))
  ),
  tibble(
    Link      = "identity",
    Pathogen  = c("RSV", "Influenza"),
    Burden    = c(sum(id_m_comp_rsv), sum(id_m_comp_flu)),
    Total_fit = sum(fitted(fit_id_m))
  )
) |>
  mutate(
    Pct       = Burden / Total_fit * 100,
    bar_label = sprintf("%s  (%.1f%%)", comma(round(Burden)), Pct)
  )

knitr::kable(
  burden_m |> select(Link, Pathogen, Burden, Pct),
  digits    = c(0, 0, 0, 1),
  caption   = "Table 4: Multiplicative DGP — 10-year attributable burden by link function",
  col.names = c("Link", "Pathogen", "Attributable hospitalizations", "% of total fitted")
)
```

```{r mult-burden-plot}
#| fig-cap: "Multiplicative DGP: 10-year attributable burden from the correctly-specified log-link model and the misspecified identity-link model. The dashed line marks the true burden derived from the data-generating process. Under a multiplicative DGP the log-link model should recover the true burden more accurately."
#| fig-height: 3.5

ggplot(burden_m, aes(x = Pathogen, y = Burden, fill = Link)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  geom_text(
    aes(label = bar_label),
    position = position_dodge(width = 0.6),
    vjust    = -0.4,
    size     = 3.2,
    fontface = "bold",
    color    = "gray20"
  ) +
  geom_crossbar(
    data        = true_burden_m,
    aes(x = Pathogen, y = TrueBurden, ymin = TrueBurden, ymax = TrueBurden),
    inherit.aes = FALSE,
    width       = 0.65,
    color       = "black",
    linewidth   = 0.7,
    linetype    = "dashed",
    fatten      = 0
  ) +
  scale_fill_manual(
    values = c("log" = "#2166ac", "identity" = "#d6604d"),
    labels = c("log" = "Poisson (log)", "identity" = "Poisson (identity)")
  ) +
  scale_y_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    x        = NULL,
    y        = "Attributable hospitalizations (10-year total)",
    fill     = NULL,
    title    = "Figure 8: Attributable Burden — Multiplicative DGP",
    subtitle = "Excess above seasonal baseline, by pathogen and link function\nDashed line = true value from data-generating process"
  ) +
  theme_ts(legend.position = "top")
```
